Applied Measure Theory for Probabilistic Modeling

### 2022-05-12 paper/paper.tex
edits

\usepackage{inconsolata}
\usepackage[normalem]{ulem} %\sout
\usepackage{centernot} %\centernot
%\usepackage[colorinlistoftodos]{todonotes}
\usepackage{hyperref}
\setcounter{page}{1}
\include{moritzdefs}
\setlength{\parskip}{0.5em}
\newcommand{\supp}{\operatorname{supp}}
\newcommand{\chad}[2][] {\todo[inline,backgroundcolor=orange!20!white, size=\footnotesize,#1]{(Chad) #2 }}
\newcommand{\moritz}[2][] {\todo[inline,backgroundcolor=blue!10!white, size=\footnotesize,#1]{(Moritz) #2 }}
\begin{document}

@@

Note the difference from a typical distribution-oriented implementation; we have changed the base measure in order to simplify the log-density. This leads to better compositional efficiency.

@@

Calls to \verb|rand|, \verb|logdensity|, etc then delegate to methods according to the appropriate names. We use \verb|KeywordCalls.jl| \cite{KeywordCalls.jl}, so all names are resolved statically at compile time.

@@

A \emph{kernel} is a (measurable\footnote{That is, the function $x \mapsto \kappa(x)(A)$ must be measurable for every fixed $A$.}) function $\kappa$ that returns a measure. Equivalently, it represents a family of measures parameterized by its argument. Writing $\cM(Y)$ for ``measures on $Y$'', we can write this as
\[
\kappa\colon X \to \mathscr{M}(Y)\ .
\]

A prominent application is a \href{https://en.wikipedia.org/wiki/Conditional_probability_distribution}{\emph{conditional distribution}}. In this case $\kappa$ is further restricted to be a \href{https://en.wikipedia.org/wiki/Markov_kernel}{\emph{Markov kernel}}, 
\[
\kappa\colon X \to \mathscr{P}(Y)\ ,
\]
where $\mathscr{P}(Y)$ represents ``probability measures on $Y$''.
If $\bX$ and $\bY$ are random variables defined on $X$ and $Y$, the kernel $\kappa$ defines a probability measure assigning every measurable $B\subset Y$ probability
\[
P(\bY \in B \mid \bX = x) =: (\kappa(x))(B)\ .
\]

@@

A Markov kernel, in particular, corresponds to the distribution of a \emph{parametrized random variable}, a function with random outcomes (or to ``mechanisms'' in causal inference, for example in \cite{https://doi.org/10.5281/zenodo.1005091}). For example in \verb|MeasureTheory|, the kernel
\[
\kappa\colon x \mapsto \cN(x,\sqrt{x})
\]
can be expressed as
\begin{verbatim}
    κ = kernel(Normal) do x
        (μ=x, σ=√x)
    end
\end{verbatim}
or equivalently
\begin{verbatim}
    κ = kernel(Normal; μ=identity, σ=sqrt)
\end{verbatim}
corresponding to the random function
\begin{verbatim}
    f(rng, x) = x + √x * randn(rng)
\end{verbatim}

@@

A second special case is when we have a collection of values together with a kernel. For this we use the \verb|For| combinator. So in a regression model, we might express the response as coming from
\begin{verbatim}
    For(1:n) do j  
        Normal(β * x[j], σ)
    end
\end{verbatim}
This is exactly the product measure
\[
\bigotimes_{j=1}^n \cN(\beta x_j,\ \sigma)\ .
\]

>>>>

\usepackage{inconsolata}
\usepackage[normalem]{ulem} %\sout
\usepackage{centernot} %\centernot
\usepackage[colorinlistoftodos]{todonotes}
\usepackage{hyperref}
\setcounter{page}{1}
\include{moritzdefs}
\setlength{\parskip}{0.5em}
\newcommand{\supp}{\operatorname{supp}}
\newcommand{\chad}[2][] {\todo[inline,backgroundcolor=orange!20!white, size=\footnotesize,#1]{(Chad) #2 }}
\newcommand{\moritz}[2][] {\todo[inline,backgroundcolor=blue!10!white, size=\footnotesize,#1]{(Moritz) #2 }}
\begin{document}

@@

Note the difference from a typical distribution-oriented implementation; we have changed the base measure in order to simplify the log-density. This leads to better compositional efficiency. \chad{"compositional efficiency" is kind of vague, reword}

The way \verb|MeasureTheory.jl| is set up, repeated application of \verb|basemeasure| will eventually reach a fix point. That is, some measure will eventually be its own base measure. For such a measure, \verb|logdensity_def| is required to return zero. We refer to the fix point reached in this way from a given measure as that measure's \emph{root measure}.

\subsection*{The end-user perspective}

We could easily have chosen a different base measure and adjusted \verb|logdensity_def| accordingly.
From the perspective of an end user, this could be considered arbitrary.
Therefore, \verb|basemeasure| and \verb|logdensity_def| should be considered implementation details, and not part of the user-facing interface.

@@

Calls to \verb|rand|, \verb|logdensity_def|, etc then delegate to methods according to the appropriate names. We use \verb|KeywordCalls.jl| \cite{KeywordCalls.jl}, so all names are resolved statically at compile time.

@@

A \emph{kernel} is a (measurable\footnote{That is, the function $x \mapsto \kappa(x)(A)$ must be measurable for every fixed $A$.}) function $\kappa$ that returns a measure. Equivalently, it represents a family of measures parameterized by its argument. Writing $\cM(Y)$ for ``measures on $Y$'', we can write this as
\[
\kappa\colon X \to \cM(Y)\ .
\]

A prominent application is a \href{https://en.wikipedia.org/wiki/Conditional_probability_distribution}{\emph{conditional distribution}}. In this case $\kappa$ is further restricted to be a \href{https://en.wikipedia.org/wiki/Markov_kernel}{\emph{Markov kernel}}, 
\[
\kappa\colon X \to \cP(Y)\ ,
\]
where $\cP(Y)$ represents ``probability measures on $Y$''.
If $\bX$ and $\bY$ are random variables defined on $X$ and $Y$, the kernel $\kappa$ defines a probability measure assigning every measurable $B\subset Y$ probability
\[
P(\bY \in B \mid \bX = x) =: (\kappa(x))(B)\ .
\]

@@

A Markov kernel, in particular, corresponds to the distribution of a \emph{parametrized random variable}, a function with random outcomes (or to ``mechanisms'' in causal inference, for example in \cite{https://doi.org/10.5281/zenodo.1005091}). For example in \verb|MeasureTheory|, the kernel
\[
\kappa\colon x \mapsto \text{Normal}(x,\sqrt{x})
\]
can be expressed as
\begin{verbatim}
    κ = kernel(Normal) do x
        (μ=x, σ=√x)
    end
\end{verbatim}
or equivalently
\begin{verbatim}
    κ = kernel(Normal; μ=identity, σ=sqrt)
\end{verbatim}
corresponding to the random function
\begin{verbatim}
    f(rng, x) = x + √x * randn(rng)
\end{verbatim}

@@

A second special case is when we have a collection of values together with a kernel. For this we use the \verb|For| combinator. So in a regression model, we might express the response as coming from
\begin{verbatim}
    For(1:n) do j  
        Normal(β * x[j], σ)
    end
\end{verbatim}
This is exactly the product measure
\[
\bigotimes_{j=1}^n \text{Normal}(\beta x_j,\ \sigma)\ .
\]
